---
title: "Bolas_Geo200CN_Final"
author: "Ellie Bolas,Geo200CN, Spring 2017"
date: "June 1, 2017"
output: html_document
---
#Final Project
Using half of the photos from a single grid, SC04-Pozo to test out occupancy modeling and how to use the camtrapR package

####read in data: from MapView
```{r}
#install.packages("camtrapR")
library(camtrapR)
library(dplyr)
library(tidyr)
#install.packages("secr")
library(secr)
#install.packages("unmarked")
library(unmarked)

pozo12a <- read.csv("data/SC04012_20161214.csv", header = TRUE)
head(pozo12a)

pozo6a <- read.csv("data/SC04006_20161214.csv", header = TRUE)
head(pozo6a)

pozoMa <- read.csv("data/SC0400M_20161214.csv", header = TRUE)

```


####clean data: from MapView
I want to do two things to get the data clean:
1. Remove the un-needed columns
2. Make a new column that combines label and Image.Name. That way the location, date, and time are all together and large amounts of data can be manipulated without getting confused.
3. Add utms (in Nad83)
4. Seperate out the tags
```{r}
#remove the columns we don't need
pozo12 <- pozo12a %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

pozo6 <- pozo6a %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

pozoM <- pozoMa %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

#new column combining label and Image.Name
pozo12$imagename <- paste(pozo12$Label, "_", pozo12$Image.Name)

pozo6$imagename <- paste(pozo6$Label, "_", pozo6$Image.Name)

pozoM$imagename <- paste(pozoM$Label, "_", pozoM$Image.Name)

#UTMS column
pozo12$utmx <- 235842
pozo12$utmy <-  3766396

pozo6$utmx <- 236156
pozo6$utmy <- 3765434

pozoM$utmx <- 235995
pozoM$utmy <- 3765908

#separate tags
#R sees the data as Target\nIsland Fox, \n is a line break, which is an escape character. "\" or "\n" is a character itself

#Check what the separators look like when there is more than two tags
#pozo12[100,] Target\nIsland Spotted Skunk\ninteresting
#I think this doesn't matter. All I care about right now are the first two

class(pozoM) #data.frame
head(pozoM)
#trying the separate function..
pozo12sep <-separate(data = pozo12, col = Tags, into = c("targettag", "speciestag"), sep = "\n", remove = FALSE, fill = "right")

pozo6sep <- separate(data = pozo6, col = Tags, into = c("targettag", "speciestag"), sep = "\n", remove = FALSE, fill = "right")

pozoMsep<- separate(data = pozoM, col = Tags, into = c("targettag", "speciestag"), sep = "\n", remove = FALSE, fill = "right")

```

##Notes on camtrap R
CamtrapR is a workflow and set of operations to take a camera trapper from raw photos to some exploratory stats and data ready to be analysed. The features can help prep for occupancy modeling or spatial-recapture analysis, the two common ways of using camera data. Island foxes and island spotted skunks are not distinguishable to individual, so I will use occupany analysis to answer my varied questions. Many of the functions camtrapR offers, I already am able to do through my tagging software, MapView. So, the questions are:
-can what I have done translate to camtrapR? Do I need it to? -Or, can I immediately jump into using secr or unmarked or presence?

###camtrapR components
####-Image management and species identifacation: 
Starts with ideas about how to organize and and tag photos, I already have a system I'm happy with

####-Image Data Extraction
  -record table: I already have this. BUT, when camtrapR does a record table, it does the "target" for me, by calculating time between images. I should test this.
  -Camera Station Operation table: The reflects the daily trapping effort per station. We need this info, both because I need to be able to calculate this information, and also because camtrapR uses this to do other things. I also need to make this, camtrapR recommends making it in excel and then importing it. 
  
#####recordTable 
```{r}
#pozo6record<- recordTable(inDir = pozo6sep
#                          IDfrom = "metadata"
#                         minDeltaTime = 1 #in minutes
#                          deltaTimeCompared to = "last record"
#)
#Ok, this isn't working because I already have data in a CSV, it will only work if I do it with photos in some kind of directory. I already have my interns do this. But maybe I should use this program?????

```

#####cameraOperation
```{r}
SC04tbl<-read.csv("data/SC04StationInfo.csv")
SC04operat<- cameraOperation(CTtable = SC04tbl,
                stationCol = "Station",
                setupCol = "setup_date",
                retrievalCol = "retrieval_date",
                writecsv = FALSE,
                hasProblems = FALSE,
                dateFormat = "%d/%m/%Y"
)
#head(SC04operat)

```

####-Data Visualization and Exploration
  -detectionMaps: plots maps of species records by staton and number of independent detections by species
  -single-species activity patterns: histograms of hourly activity, activity kernal density estimations, radial plots (activityHistogram, activityDensity, activityRadial)
  -two-species activity overlaps: activityOverlap
Playing with this stuff will be one way to assess whether my tagging system/recordtable is fine, and/or how I need to modify this part of things. 

#####detectionMaps
For detectionMaps, the station column should be a station ID that is the same in both the CTtable and recordTable. (Maybe this is how the two )
```{r}
#pozo12map <- detectionMaps(CtTable = SC04operat, 
#                           recordTable = pozo12sep,
#                           Xcol = "utm_x",
#                           Ycol = "utm_y",
#                           stationCol = "station"
#)


```



Other things to do: 
-test digikam, maybe it's better than mapview
-get to know secr package and capthist function
-find a way to inclue total days cam running

