---
title: Bolas_Lab_2
author: Ellie Bolas, GEO200CN, Spring 2016
output: html_document
---
```{r set up and load packages}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../"))

library(readr)
library(dplyr)

```

#Pointers from Elise
```{r Pointers from Elise}
a<- 1:5
b<- 6:10
a*b #multiplies element-wise
c<- 1:10
a+c #"c" is larger, when it runs out of elements from "a," it starts back over at the beginning of "a"

```


__Carefully read chapter 7 to 14 in http://rspatial.org/intr/ __


#x__1) Functions+Apply__

a) Use the files from lab 1 to compute mean values for each variable using 'apply'

```{r read in data}
datapath <- "../GEO200CN/data"


lob1<- read.table(file.path(datapath, "loblolly.dat"), header = TRUE) 

#file.path uses the complete path only to its folder


chick1<- read.csv(file.path(datapath, "ChickWeightSub.csv"))

```

```{r means}
apply(lob1, 2, mean)

apply(chick1, 2, mean)
```


b) Write a function 'f' that behaves like this:

  > f('Sarah')
  [1] "hello Sarah, how are you"
  
  
```{r name function}
f<-function(name) {
  x<- paste("hello ", name,", how are you", sep = "")
  return(x)
}

f("Sarah")
#paste0() seperator is set to be nothing

d<-function(name) {
   x<- paste0("hello ", name, ",", " how are you")
   return(x)
}
     
d("Ellie")
  
```

c) Write a function 'sumOfSquares' that behaves like this:

  > d <- c(1,5,2,4,6,2,4,5)
  > sumOfSquares(d)
  [1] 21.875

To compute the "sum of squares", subtract the mean value of all numbers from each number. Square these numbers and sum them

```{r sumofSquares function}
d <- c(1,5,2,4,6,2,4,5)
d

sumofSquares <- function(d) {
  meand <- mean(d)
  submean <- d-meand
  sqsum <- sum(submean*submean)
  return(sqsum)
}

sumofSquares(d)

```


__2) Data preparation+exploration__. Use the data in "students.csv" and "scores.dat" to assess if there is a relation between the "group" or "age" of a student and their average score and for the three subjects (math, english, geography)

just correlate between average and group and average and age
```{r read in data and load packages}
scores <- read.table(file.path(datapath, "scores.dat"), header = TRUE)

students1 <- read.csv(file.path(datapath, "students.csv"))

library(dplyr)
library(plyr)

```

a) use reshape and merge to combine the two data sets
```{r combine data sets}
scores1 <- rename(scores, c("student" = "students"))

students_joined <-full_join(students1, scores1, by = NULL)
  

```



b) compute correlation coefficients
c) produce a relevant plot


__Hand in the results for #1 and #2 as a R Markdown file.__ Both the source file and an HTML file. You can complete the .Rmd file available on smartsite.
 

