---
title: "Lab 1, Ellie Bolas"
author: "Ellie Bolas, GEO200CN, Spring 2016"
output:
  html_document: default
  pdf_document: default
---
```{r set up and load packages}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../"))

library(readr)
library(dplyr)

```



_x_1) Carefully read chapter 1 to 6 in http://rspatial.org/intr/ __


_x_2) Read and explore data __

There are two text files with data (loblolly.dat) and (ChickWeightSub.csv). 

Write a script (including brief comments) that

a) Reads them into R
#Read in Data#
```{r read in data}
datapath <- "../GEO200CN/data"


lob1<- read.table(file.path(datapath, "loblolly.dat"), header = TRUE) 

#file.path uses the complete path only to its folder


chick1<- read.csv(file.path(datapath, "ChickWeightSub.csv"))

```


b) for both data sets, lists all the variable names
#Variable Names#
```{r variable names}
head(chick1)
#variables for ChickWeightSub =  Weight DayOld ChickID DietID

head(lob1)
#variables for Loblolly =  height age Seed

```


c) for each variable in one of the data sets, use a function to compute the min and a function to compute the mean value 
#Min and Mean#
```{r min and means}
lob1 %>% summarize(min_height = min(height, na.rm = TRUE))

#  min_height
# 1       3.46

chick1 %>%  summarize (min_weight = min(Weight, na.rm =TRUE))
#  min_weight
#1         39

min(lob1$age) #[1] 3
min(lob1$Seed) #[1] 301

mean(lob1$height)
mean(lob1$age)
mean(lob1$Seed)

#apply goes through each column or each row of a matrix and applies the function to each of those

#lapply apply this function to each element of a list

apply(lob1, 2, mean) #(dataframe, 2 means each column, then funcntion)

```


d) now pick one variable from one data set and use functions summary, table, quantile, and boxplot

#One Variable Manipulation
```{r Lob1 Height}
summary(lob1$height)
table(lob1$height)
quantile(lob1$height)
#shows where the breaks in the interquantile range fall
boxplot(lob1$height)

```



x__Hand in the results as a R Markdown file.__ Both the source file and a pdf. You can complete the .Rmd file available on smartsite.


