---
title: "Bolas_Geo200CN_Final: camtrapR"
author: "Ellie Bolas,GEO200CN, Spring 2017"
date: "June 1, 2017"
output: html_document
---
#Final Project: camtrapR
Using half of the photos from a single grid, SC04-Pozo to test out how to use the camtrapR package. 

camtrapR is designed to help get raw photos into a format ready to be analyzed in unmarked, a package that does occupancy modeling. camtrapR also allows data visualization and exploration, by making species richness and detection maps. It can also return activity patterns.

####read in data: from MapView
```{r}
#install.packages("camtrapR")
library(camtrapR)
library(dplyr)
library(tidyr)
#install.packages("secr")
library(secr)
#install.packages("unmarked")
library(unmarked)


pozo12a <- read.csv("data/SC04012_20161214.csv", header = TRUE)
head(pozo12a)

pozo6a <- read.csv("data/SC04006_20161214.csv", header = TRUE)
head(pozo6a)

pozoMa <- read.csv("data/SC0400M_20161214.csv", header = TRUE)

```


####clean data: from MapView
I want to do two things to get the data clean:
1. Remove the un-needed columns
2. Make a new column that combines label and Image.Name. That way the location, date, and time are all together and large amounts of data can be manipulated without getting confused.
3. Add utms (in Nad83)
4. Seperate out the tags
```{r}
#remove the columns we don't need
pozo12 <- pozo12a %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

pozo6 <- pozo6a %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

pozoM <- pozoMa %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

#new column combining label and Image.Name
pozo12$imagename <- paste(pozo12$Label, "_", pozo12$Image.Name)

pozo6$imagename <- paste(pozo6$Label, "_", pozo6$Image.Name)

pozoM$imagename <- paste(pozoM$Label, "_", pozoM$Image.Name)

#UTMS column
pozo12$utmx <- 235842
pozo12$utmy <-  3766396

pozo6$utmx <- 236156
pozo6$utmy <- 3765434

pozoM$utmx <- 235995
pozoM$utmy <- 3765908

#separate tags
#R sees the data as Target\nIsland Fox, \n is a line break, which is an escape character. "\" or "\n" is a character itself

#Check what the separators look like when there is more than two tags
#pozo12[100,] Target\nIsland Spotted Skunk\ninteresting
#I think this doesn't matter. All I care about right now are the first two

class(pozoM) #data.frame
head(pozoM)
#seperate out all the tags, which are currently in a single column
pozo12sep <-separate(data = pozo12, col = Tags, into = c("targettag", "speciestag"), sep = "\n", remove = FALSE, fill = "right")

pozo6sep <- separate(data = pozo6, col = Tags, into = c("targettag", "speciestag"), sep = "\n", remove = FALSE, fill = "right")

pozoMsep<- separate(data = pozoM, col = Tags, into = c("targettag", "speciestag"), sep = "\n", remove = FALSE, fill = "right")

```

##Notes on camtrap R
CamtrapR is a workflow and set of operations to take a camera trapper from raw photos to some exploratory stats and data ready to be analysed. The features can help prep for occupancy modeling or spatial-recapture analysis, the two common ways of using camera data. Island foxes and island spotted skunks are not distinguishable to individual, so I will use occupany analysis to answer my varied questions. Many of the functions camtrapR offers, I already am able to do through my tagging software, MapView. So, the questions are:
-can what I have done translate to camtrapR? Do I need it to? -Or, can I immediately jump into using secr or unmarked or presence?

##camtrapR components
###-Image management and species identifacation: 
Starts with ideas about how to organize and and tag photos, I already have a system I'm happy with

###-Image Data Extraction
  -record table: I already have this. BUT, when camtrapR does a record table, it does the "target" for me, by calculating time between images. I should test this.
  -Camera Station Operation table: The reflects the daily trapping effort per station. We need this info, both because I need to be able to calculate this information, and also because camtrapR uses this to do other things. I also need to make this, camtrapR recommends making it in excel and then importing it. 
  
#####recordTable 
```{r}
#pozo6record<- recordTable(inDir = pozo6sep
#                          IDfrom = "metadata"
#                         minDeltaTime = 1 #in minutes
#                          deltaTimeCompared to = "last record"
#)
#Ok, this isn't working because I already have data in a CSV, it will only work if I do it with photos in some kind of directory. I already have my interns do this. But maybe I should use this program?????

```

####Turning my current table into a recordTable like camtrapR
Their record table has station, species, date, time, and this delta.time, which is a way to identify a target. I will make a record table with similar information.


```{r}
#select is for columns, filter is for rows

#rt12 <- pozo12sep %>% select(Label, Date, Time, targettag, speciestag, imagename) %>%  filter(!is.na(speciestag)& !(speciestag== "interesting"))
#this worked! Got rid of the column fill-in problem: some photos have more than the two basic tags of "target" and "species name," I had to figure out how to not include that in this analysis Double-checked this against photos in mapview, 37 "targets" in both locations

rtp12 <- pozo12sep %>% select(Label, Date.Time, Date, Time, targettag, speciestag, imagename) %>%  filter(!is.na(speciestag)& !(speciestag== "interesting") & !(speciestag == "Other") & !(speciestag == "Wind/Heat")) #this now only includes animal species

rtp6 <-pozo6sep %>% select(Label, Date.Time, Date, Time, targettag, speciestag, imagename) %>%  filter(!is.na(speciestag)& !(speciestag== "interesting") & !(speciestag == "Other") & !(speciestag == "Wind/Heat"))

rtpM <- pozoMsep %>% select(Label, Date.Time, Date, Time, targettag, speciestag, imagename) %>%  filter(!is.na(speciestag)& !(speciestag== "interesting") & !(speciestag == "Other") & !(speciestag == "Wind/Heat"))

#record table for Pozo!
recordTablePozo <- bind_rows(rtp12, rtp6, rtpM)
summary(recordTablePozo)

#need to add a DateTimeOriginal column
#Did this in excel. Ultimately, better to do this in R, but it's a little clunky to make work in R right now. Will re-visit later.


```


#####cameraOperation
```{r}
SC04tbl<-read.csv("data/SC04StationInfo.csv")


#it's important to make sure that the date format for setup_date and retrieval_date are the same format as the Date.Time column in the record table
SC04operat<- cameraOperation(CTtable = SC04tbl,
                stationCol = "Label",
                setupCol = "setup_date",
                retrievalCol = "retrieval_date",
                writecsv = FALSE,
                hasProblems = FALSE,
                dateFormat = "%Y-%m-%d"
)
head(SC04operat)

```

###-Data Visualization and Exploration
  -detectionMaps: plots maps of species records by staton and number of independent detections by species
  -single-species activity patterns: histograms of hourly activity, activity kernal density estimations, radial plots (activityHistogram, activityDensity, activityRadial)
  -two-species activity overlaps: activityOverlap
Playing with this stuff will be one way to assess whether my tagging system/recordtable is fine, and/or how I need to modify this part of things. 

#####detectionMaps
For detectionMaps, the station column should be a station ID that is the same in both the CTtable and recordTable. (Maybe this is how the two merge)
```{r}
pozomap <- detectionMaps(CTtable = SC04tbl, 
                          recordTable = recordTablePozo,
                          Xcol = "utm_x",
                          Ycol = "utm_y",
                          stationCol = "Label",
                         speciesCol = "speciestag",
                         writePNG = FALSE,
                         plotR = TRUE,
                         richnessPlot = TRUE,
                         speciesPlots = TRUE,
                         printLabels = TRUE,
                         addLegend = TRUE
)

```
This printed out a species richness plot, as well as a plot for detections for each "species." I'll need to fix "bird" and "mouse." These plots are great, they will be even more exciting when I have every grid on here. 

#####Activity Patterns
```{r}
#diel activity histograms
allactivityhist<- activityHistogram(recordTable = recordTablePozo,
allSpecies = TRUE,
speciesCol = "speciestag",
recordDateTimeCol = "Date.Time",
recordDateTimeFormat = "%Y-%m-%d %H-%M-%S",
plotR= TRUE)
#These histograms are awesome! With more data to analyze, there will be some nice things to see about time of day activity. Foxes are actually active at all times of day, and a bigger data set will reflect that.

#kernal density overlaps
foxskunkoverlap <- activityOverlap(recordTable = recordTablePozo, 
speciesA = "Island Fox",
speciesB = "Island Spotted Skunk",
speciesCol = "speciestag",
recordDateTimeCol = "Date.Time",
recordDateTimeFormat = "%Y-%m-%d %H-%M-%S",
plotR= TRUE,
linecol = c("red", "blue"), 
linewidth = c(3,3))

```


###-Data Export for Occupancy Analysis
The unmarked package uses a custom df for analyses. camtrapR claims to be the best system to get ready for occupancy analyses (which are done with either the unmarked package, or other programs/pacakges.) I have only looked over unmarked a little bit at this point, so I'm not totally clear on how much of camtrapR really is getting me where I need to go. But, this is the last step, so I want to test this capability and what it gives me

Single Species detectionHistory outputs 
```{r}
dhpozofox <- detectionHistory(recordTable = recordTablePozo,
                             camOp = SC04operat,
                           species = "Island Fox",  
                           stationCol = "Label",
                           speciesCol = "speciestag",
                           recordDateTimeCol = "Date.Time",
                           recordDateTimeFormat = "%Y-%m-%d %H-%M-%S",
                           occasionLength = 1,#in days
                           day1 = "2016-08-25", #means the day the station was set
                           includeEffort = TRUE,
                           scaleEffort = FALSE,
                           occasionStartTime = 6, #default is midnight, that throws things off for skunks/nocturnal animals
                           timeZone = "US/Pacific",
                           writecsv = TRUE,
                           outDir = "C:/Users/ebola/Google Drive/Git/GEO200CN/data"
)
dhpozofox[[1]] #detection history
dhpozofox[[2]] #effort

dhpozoskunk <- detectionHistory(recordTable = recordTablePozo,
                             camOp = SC04operat,
                           species = "Island Spotted Skunk",  
                           stationCol = "Label",
                           speciesCol = "speciestag",
                           recordDateTimeCol = "Date.Time",
                           recordDateTimeFormat = "%Y-%m-%d %H-%M-%S",
                           occasionLength = 1,#in days
                           day1 = "2016-08-25", #means the day the station was set
                           includeEffort = TRUE,
                           scaleEffort = FALSE,
                           occasionStartTime = 6, #default is midnight, that throws things off for skunks/nocturnal animals
                           timeZone = "US/Pacific",
                           writecsv = TRUE,
                           outDir = "C:/Users/ebola/Google Drive/Git/GEO200CN/data"
)
dhpozoskunk[[1]]

```
What I have learned from this is that in using mapview, I was still able to coerce the data to give me the outputs the camtrapR offers. I don't need to add metadata to exif tables to do this, nor do I have to make a bunch of seperate species folders per camera station. It also gives me the ability to manipulate the descritization of the time period. That is, when I tag "target" on my own, I can find a way to note multiple species, or get the timing right on what is a seperate individual to the trap.

##-Summary
```{r}
#surveyReport(recordTable = recordTablePozo,
#             CTtable = SC04tbl,
#             stationCol = "Label",
#            speciesCol = "speciestag",
#            recordDateTimeCol = "Date.Time",
#            recordDateTimeFormat = "%Y-%m-%d %H-%M-%S",
#            setupCol = "setup_date",
#            retrievalCol = "retrieval_date",
#            CTDateFormat = "%Y-%m-%d", 
#             CTHasProblems = FALSE,
#            Xcol = "utm_x",
#            Ycol = "utm_y",
#            sinkpath = "C:/Users/ebola/Google Drive/Git/GEO200CN",
#              makezip = TRUE)
#Didn't make the entire zipfile, not sure why, but the textfile is cool
```








