---
title: "Bolas_Geo200CN_Final"
author: "Ellie Bolas,Geo200CN, Spring 2017"
date: "June 1, 2017"
output: html_document
---
#Final Project
Using half of the photos from a single grid, SC04-Pozo to test out occupancy modeling and how to use the camtrapR package

####read in data
```{r}
#install.packages("camtrapR")
library(camtrapR)
library(dplyr)
#datapath <- "C:/Users/ebola/Google Drive/Git/GEO200CN/data"
pozo12a <- read.csv("data/SC04012_20161214.csv", header = TRUE)
head(pozo12a)

pozo6a <- read.csv("data/SC04006_20161214.csv", header = TRUE)
head(pozo6a)

pozoMa <- read.csv("data/SC04006_20161214.csv", header = TRUE)

```


####clean data
I want to do two things to get the data clean:
1. Remove the un-needed columns
2. Make a new column that combines label and Image.Name. That way the location, date, and time are all together and large amounts of data can be manipulated without getting confused.
3. Add utms (in Nad83)
4. Seperate out the tags
```{r}
#remove the columns we don't need
pozo12 <- pozo12a %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

pozo6 <- pozo6a %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

pozoM <- pozoMa %>% select(-Trigger, -Moon.Phase, -Illum, -Contrast, -Brightness, -Sharpness, -Sensitivity, -Battery.Volts., -Saturation, -Serial.., -Narrative, -Firmware, -Image.Path, -Ir, -Vir, -Original.File.Name, -Location.Latitude, -Location.Longitude)

#new column combining label and Image.Name
pozo12$imagename <- paste(pozo12$Label, "_", pozo12$Image.Name)

pozo6$imagename <- paste(pozo6$Label, "_", pozo6$Image.Name)

pozoM$imagename <- paste(pozoM$Label, "_", pozoM$Image.Name)

#UTMS column
pozo12$utmx <- 235842
pozo12$utmy <-  3766396

pozo6$utmx <- 236156
pozo6$utmy <- 3765434

pozoM$utmx <- 235995
pozoM$utmy <- 3765908
```


